Using spatially biased variables in ecosystem condition accounting with a GIS based workflow

### 2024-02-21 manuscript/HIAs.qmd
fix methods map

library(tidyverse)
library(knitr)
library(sf)
library(tmap)
library(stars)
library(terra)
library(tidyterra)
library(ggtext)
library(cowplot)
library(units)

@@

# Mire data
mire_terra <- terra::rast(path_mire)
```

@@

```{r positionMap}
#| eval: false

inc <- 1e5
myBbox <- st_bbox(centroids)
myBbox[1:2] <- myBbox[1:2]-inc 
myBbox[3:4] <- myBbox[3:4]+inc

(positionMap <- 
  tm_shape(outline,
           bbox = myBbox) +
  tm_polygons() +
  tm_shape(centroids) +
  tm_text(
    text = "Municipality",
    just= "left",
    size = .8,
    xmod = 1,
    ymod = 0
  ) +
  tm_symbols(
    size = 1.5,
    shape = 17,
    col = "black"
  ) +
  tm_grid(projection = 4326) +
  tm_layout(
    bg.color = "skyblue",
    outer.margins = c(0.01, .02, .02, .02))+
  tm_compass()+
  tm_scale_bar()+
    tm_basemap(leaflet::providers$Stamen.Watercolor)
)

@@

# what is the distance between Nordre Follo and Målselv
(latlong <- centroids |>
  st_transform(4326) |>
st_coordinates() |>
  as_tibble())

km_distance <- centroids |>
  st_distance() |>
  max() |>
  set_units("km") |>
  drop_units() |>
  round()
#round(max(latlong$Y)-min(latlong$Y), 1)
```

```{r mireDataTestCrop}
#| eval: false

# Testing the cropping proceedure using a downsampled starts object
mire_test <- read_stars(mireFile, proxy = T) |>
  setNames("mire") |>
  st_downsample(100) |>
  mutate(mire = na_if(mire, 0)) |>
  st_as_stars()

mire_test_nf <- mire_test |>
  st_crop(nf)

# The cropping is correct:
# tmap_mode("plot")
# tm_shape(mire_test)+
#  tm_raster(palette = "red")+
#  tm_shape(outline)+
#  tm_borders()+
#  tm_shape(mire_test_nf)+
#  tm_raster(col="red")

# Then we do the same, but without downsampling

mire_nf <- read_stars(mireFile, proxy = T) |>
  setNames("mire") |>
  # mutate(mire = na_if(mire, 0)) |>
  st_crop(nf) |>
  st_as_stars()
# The cropped raster is for some reason far into the ocean!

# The cells have values, so it seems like there is something
# that happens after the cropping
tmap_mode("plot")
tm_shape(mire_test) +
  tm_raster(palette = "red") +
  tm_shape(outline) +
  tm_borders() +
  tm_shape(mire_nf) +
  tm_raster(col = "red")
```

```{r mireTerra}
#| eval: false

# trying { terra } instead

# convert municipal outline to vect via st
nf_vect <- as(nf, "Spatial") |>
  terra::vect()

# crop and mask (very fast!)
mire_nf_terra <- terra::rast(mireFile) |>
  terra::crop(nf_vect) |>
  terra::mask(nf_vect)

@@

# I like the stars, sf and tmap combo better, so I return to stars
mire_nf_stars <- mire_nf_terra |>
  st_as_stars()

ggplot() +
  geom_stars(data = mire_nf_stars) +
  coord_equal() +
  theme_void() +
  scale_fill_viridis_c()
```

@@

# calculate area of survey coverage maps
# values goes into summary table in the ms
dk2 <- read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |>
  st_transform(st_crs(muni3)) |>
  st_intersection(muni3) |>
  st_cast("POLYGON") |>
  group_by(Municipality) |>
  summarise(geometry = st_union(geometry)) |>
  mutate(
    dk_area_km = geometry |> st_area(),
    dk_area_km = round(units::drop_units(dk_area_km * 1e-6))
  )
```

@@

# calculate area of the mires in each municipality
# -- Nordre Follo
mireArea <- mire_nf_terra |>
  global(c("mean", "sum"), na.rm = T) |>
  add_column("Municipality" = "Nordre Follo") |>
  mutate(
    mirePercent = round(mean * 100, 1),
    mire_km2 = sum / 1e+4
  )

# Calculate the area of mire inside the coverage maps
# -- Nordre Follo
mire_nf_terra_dk <- mire_nf_terra |>
  terra::mask(dk2 |> filter(Municipality == "Nordre Follo")) |>
  global("sum", na.rm = T) |>
  mutate(mireInSurvey_km2 = sum / 1e+4) |>
  add_column(Municipality = "Nordre Follo")
```

```{r infrastructureIndex}
#| eval: false

# this data is on a 100x100m grid
infra <- stars::read_stars(path_infrastructure) |>
  # , which is mor then we need - warp it to 1x1km
  st_warp(
    cellsize = c(1000, 1000),
    crs = st_crs(nf),
    use_gdal = TRUE,
    method = "average"
  ) |>
  setNames("infrastructureIndex") |>
  st_transform(myCRS) |>
  mutate(infrastructureIndex = case_when(
    infrastructureIndex < 1 ~ 0,
    infrastructureIndex < 6 ~ 1,
    infrastructureIndex < 12 ~ 2,
    infrastructureIndex >= 12 ~ 3
  )) |>
  # takeing away point in the sea
  st_crop(outline)

# This step might seem rather stupid. We want to vectorize a rather large raster. This makes it a quite big data object. The reason is that there is no really good way to burn polygon data on to raster grid cells after the disuse of the raster package. It was not straight forward then either. But calculating intersections between polygons is very fast and easy.

@@

```{r readInfraStructureIndex}
#| eval: true

@@

# Calulculate area
infra <- infra |>
  mutate(area = geometry |> st_area(),
         area_km = area |> set_units("km2"))

# show the summered area per HIA
infra |>
  as_tibble() |>
  group_by(infrastructureIndex) |>
  summarise(area_km = sum(area_km)) |>
  ggplot(aes(x = infrastructureIndex,
             y = area_km)) +
  geom_col()

@@

# Make a plot to check that it has worked
infraMuni3_tbl |>
  ggplot() +
  geom_bar(
    aes(
      x = infrastructureIndex,
      y = total_area_HIAs_km2,
      fill = factor(infrastructureIndex),
      colour = factor(infrastructureIndex)
    ),
    stat = "identity",
    lwd = 1.2
  ) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(4, "YlOrBr")) +
  scale_color_manual(values = RColorBrewer::brewer.pal(5, "YlOrBr")[-1]) +
  theme_minimal_hgrid() +
  labs(
    x = "Human Impact Area (HIA) class",
    y = "Area (km<sup>2</sup>)"
  ) +
  theme(
    axis.title.x = element_textbox_simple(
      width = NULL,
      padding = margin(4, 4, 4, 4),
      margin = margin(4, 0, 0, 0),
      linetype = 1,
      r = grid::unit(8, "pt"),
      fill = "azure1"
    ),
    axis.title.y = element_textbox_simple(
      width = NULL,
      padding = margin(4, 4, 4, 4),
      margin = margin(4, 0, 0, 0),
      linetype = 1,
      orientation = "left-rotated",
      r = grid::unit(8, "pt"),
      fill = "azure1"
    ),
    strip.background = element_blank(),
    strip.text = element_textbox(
      size = 12,
      color = "white", fill = "#5D729D", box.color = "#4A618C",
      halign = 0.5, linetype = 1, r = unit(5, "pt"), width = unit(1, "npc"),
      padding = margin(2, 0, 1, 0), margin = margin(3, 3, 3, 3)
    )
  ) +
  guides(fill = "none", colour = "none") +
  scale_y_log10() +
  facet_grid(cols = vars(Municipality)) 
```

```{r HIA-validate}
#| eval: false

# Now I want to see if the indicator values depend on the HIA is a predictable
# way to justify the stratification
corrCheck <- st_intersection(naturetypes, infra)

#corrCheck |>
#  as_tibble() |>
#  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
#         i_alien_fct = floor(round(i_alien*10,2))/10) |>
#  count(i_ADSV_fct, infrastructureIndex) |>
#  View()

# A first look
corrCheck |>
  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
         i_alien_fct = floor(round(i_alien*10,2))/10,
         i_ditch_fct = floor(round(i_ditch*10,2))/10) |>
  pivot_longer(cols = c(i_ADSV_fct, i_alien_fct, i_ditch_fct),
               values_to = "indicatorValue",
               names_to = "indicator",
               values_drop_na =T) |>
  ggplot(aes(
    x = factor(infrastructureIndex), 
    fill = factor(indicatorValue)))+
  geom_bar(
    position="fill"
    )+
  theme_bw(base_size = 12)+
  guides(fill = guide_legend("Scaled indicator values"))+
  ylab("Fraction of data points")+
  xlab("HIA")+
  scale_fill_brewer(palette = "RdYlGn")+
  facet_grid(indicator~year)
```

# The year effect is constant, so ignoring that
corrCheck |>
  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
         i_alien_fct = floor(round(i_alien*10,2))/10,
         i_ditch_fct = floor(round(i_ditch*10,2))/10) |>
  pivot_longer(cols = c(i_ADSV_fct, i_alien_fct, i_ditch_fct),
               values_to = "indicatorValue",
               names_to = "indicator",
               values_drop_na =T) |>
  ggplot(aes(
    x = factor(infrastructureIndex), 
    fill = factor(indicatorValue)))+
  geom_bar(
    position="fill"
    )+
  theme_bw(base_size = 12)+
  guides(fill = guide_legend("Scaled indicator values"))+
  ylab("Fraction of data points")+
  xlab("HIA")+
  scale_fill_brewer(palette = "RdYlGn")+
  facet_grid(~indicator)
```

# Using a colour gradient emphasises the first colour (dark green).
# Lets try discrete colours, and merge som classes to simplify
corrCheck |>
  pivot_longer(cols = c(i_ADSV, i_alien, i_ditch),
               values_to = "indicatorValue",
               names_to = "indicator",
               values_drop_na = T) |>
  mutate(condition = case_when(
    indicatorValue <0.6 ~"Poor",
    indicatorValue <0.8 ~"Intermediate",
    indicatorValue <0.91 ~"Good",
    .default = "Perfect"
  ),
  condition = fct_reorder(condition, indicatorValue)
  ) |>
  ggplot(aes(
    x = factor(infrastructureIndex), 
    fill = condition))+
  geom_bar(
    position="fill"
    )+
  theme_bw(base_size = 12)+
  guides(fill = guide_legend("Scaled indicator values"))+
  ylab("Fraction of data points")+
  xlab("HIA")+
  scale_fill_viridis_d(alpha = .8) +
  facet_grid(~indicator)

@@

# The figure changes a lot depending on which value gets the darker colour
corrCheck |>
  pivot_longer(cols = c(i_ADSV, i_alien, i_ditch),
               values_to = "indicatorValue",
               names_to = "indicator",
               values_drop_na = T) |>
  mutate(condition = case_when(
    indicatorValue <0.6 ~"Poor",
    indicatorValue <0.8 ~"Intermediate",
    indicatorValue <0.91 ~"Good",
    .default = "Perfect"
  ),
  condition = fct_reorder(condition, -indicatorValue)
  ) |>
  ggplot(aes(
    x = factor(infrastructureIndex), 
    fill = condition))+
  geom_bar(
    position="fill"
    )+
  theme_bw(base_size = 12)+
  guides(fill = guide_legend("Scaled indicator values"))+
  ylab("Fraction of data points")+
  xlab("HIA")+
  scale_fill_viridis_d(alpha = .8) +
  facet_grid(~indicator)
```

# If I look at this on a log scale, it also changes and becomes less convincing for ADSV and alien
corrCheck |>
  pivot_longer(cols = c(i_ADSV, i_alien, i_ditch),
               values_to = "indicatorValue",
               names_to = "indicator",
               values_drop_na = T) |>
  mutate(condition = case_when(
    indicatorValue <0.6 ~"Poor",
    indicatorValue <0.8 ~"Intermediate",
    indicatorValue <0.91 ~"Good",
    .default = "Perfect"
  ),
  condition = fct_reorder(condition, -indicatorValue)
  ) |>
  ggplot(aes(
    x = factor(infrastructureIndex), 
    fill = condition))+
  geom_bar(
    position="fill"
    )+
  theme_bw(base_size = 12)+
  guides(fill = guide_legend("Scaled indicator values"))+
  ylab("Fraction of data points")+
  xlab("HIA")+
  scale_y_log10()+
  scale_fill_viridis_d(alpha = .8) +
  facet_grid(~indicator)

@@

# A figure with just the legend
infraMuniMap_l <- tm_shape(muni3) +
  tm_borders() +
  tm_shape(infraMuni3) +
  tm_polygons(
    col = "infrastructureIndex",
    style = "cat",
    title = "Human Impact\nArea Class"
  ) +
  tm_layout(legend.only = TRUE,
            legend.position = c("left", "bottom"),
            legend.outside = TRUE)

# Syrvey coverage and other colour overlayed the municipalities
muniPlot <- tm_shape(muni3) +
  tm_borders() +
  tm_facets(
    by = "Municipality",
    ncol = 3
  ) +
  tm_shape(terrestrial) +
  tm_fill(
    col = "lightgreen",
    alpha = .4
  ) +
  tm_shape(ocean) +
  tm_polygons(
    col = "skyblue",
    border.col = "black"
  ) +
  tm_shape(dk) +
  tm_polygons(
    col = "grey",
    alpha = .8
  ) +
  tm_shape(nature3) +
  tm_polygons(
    col = "red",
    border.col = "red",
    lwd = 3
  )

>>>>

library(tidyverse)
library(knitr)
library(sf)
library(tmap)
library(stars)
library(terra)
library(tidyterra)
library(ggtext)
library(cowplot)
library(units)
library(rnaturalearth)
library(rnaturalearthdata)

@@

# Mire data
mire_terra <- terra::rast(path_mire)

# Infrastructure index (read proxy)
infra <- stars::read_stars(path_infrastructure)

@@

```{r positionMap}
#| eval: true

world <- ne_countries(scale = "medium", returnclass = "sf") |>
  st_transform(myCRS) |>
  filter(admin %in% c("Norway", "Sweden")) |>
  st_make_valid()

inc <- 200000
myBbox <- st_bbox(centroids)
myBbox[1:2] <- myBbox[1:2]-inc 
myBbox[3:4] <- myBbox[3:4]+inc

(positionMap <- 
  tm_shape(world,
           bbox = myBbox) +
    tm_polygons() +
  tm_shape(muni3) +
    tm_polygons(col = "green") +
  tm_shape(centroids) +
  tm_text(
    text = "Municipality",
    just= "left",
    size = .8,
    xmod = 1,
    ymod = 0
  ) +
  tm_grid(projection = 4326) +
  tm_layout(
    bg.color = "skyblue",
    outer.margins = c(0.01, .02, .02, .02))+
  tm_compass()+
  tm_scale_bar()
)

@@

# what is the distance between Nordre Follo and Nord-Aurdal
(km_distance <- centroids |>
  st_distance() |>
  max() |>
  set_units("km") |>
  drop_units() |>
  round())
```

```{r mireTerra}
#| eval: true

# I first tried to import and crop the mire data using starts, but that failed (se pre 21 feb 2023)
# trying { terra } instead

# convert municipal outline to vect via st
nf_vect <- as(nf, "Spatial") |>
  terra::vect()
gr_vect <- as(gr, "Spatial") |>
  terra::vect()
na_vect <- as(na, "Spatial") |>
  terra::vect()

# crop and mask (very fast!)
mire_terra_nf <- mire_terra |>
  terra::crop(nf_vect) |>
  terra::mask(nf_vect)

mire_terra_gr <- mire_terra |>
  terra::crop(gr_vect) |>
  terra::mask(gr_vect)

mire_terra_na <- mire_terra |>
  terra::crop(na_vect) |>
  terra::mask(na_vect)

@@

# I like the stars, sf and tmap combo better, so I return to stars
mire_stars_nf <- mire_terra_nf |>
  st_as_stars()
mire_stars_gr <- mire_terra_gr |>
  st_as_stars()
mire_stars_na <- mire_terra_na |>
  st_as_stars()

par(mfrow=c(3,1))
plot(mire_stars_nf)
plot(mire_stars_gr)
plot(mire_stars_na)

@@

# calculate area of survey coverage maps
# values goes into summary table in the ms
dk2 <- coverage3 |>
  group_by(Municipality) |>
  summarise(SHAPE = st_union(SHAPE)) |>
  mutate(
    dk_area_km = SHAPE |> st_area(),
    dk_area_km = round(units::drop_units(dk_area_km * 1e-6))
  )
```

@@

# calculate area of the mires in each municipality
# -- Nordre Follo
(mireArea <- mire_terra_nf |>
  global(c("mean", "sum"), na.rm = T) |>
  add_column("Municipality" = "Nordre Follo") |>
  mutate(
    mirePercent = round(mean * 100, 1),
    mire_km2 = sum / 1e+4
  ))

# -- Gran
mireArea2 <- mire_terra_gr |>
  global(c("mean", "sum"), na.rm = T) |>
  add_column("Municipality" = "Gran") |>
  mutate(
    mirePercent = round(mean * 100, 1),
    mire_km2 = sum / 1e+4
  )

# -- Nord-Aurdal
mireArea3 <- mire_terra_na |>
  global(c("mean", "sum"), na.rm = T) |>
  add_column("Municipality" = "Nord-Aurdal") |>
  mutate(
    mirePercent = round(mean * 100, 1),
    mire_km2 = sum / 1e+4
  )

mireArea <- mireArea |>
  rbind(mireArea2, mireArea3)

# Calculate the area of mire inside the coverage maps
# -- Nordre Follo
mire_in_dk <- mire_terra_nf |>
  terra::mask(dk2 |> filter(Municipality == "Nordre Follo")) |>
  global("sum", na.rm = T) |>
  mutate(mireInSurvey_km2 = sum / 1e+4) |>
  add_column(Municipality = "Nordre Follo")

mire_in_dk2 <- mire_terra_gr |>
  terra::mask(dk2 |> filter(Municipality == "Gran")) |>
  global("sum", na.rm = T) |>
  mutate(mireInSurvey_km2 = sum / 1e+4) |>
  add_column(Municipality = "Gran")
mire_in_dk3 <- mire_terra_na |>
  terra::mask(dk2 |> filter(Municipality == "Nord-Aurdal")) |>
  global("sum", na.rm = T) |>
  mutate(mireInSurvey_km2 = sum / 1e+4) |>
  add_column(Municipality = "Nord-Aurdal")

mire_in_dk <- mire_in_dk |>
  rbind(mire_in_dk2, mire_in_dk3)
```

```{r infrastructureIndex}
#| eval: false
# this data is on a 100x100m grid
infra <- infra |>
  # , which is more then we need - warp it to 1x1km
  st_warp(
    cellsize = c(1000, 1000),
    crs = st_crs(nf),
    use_gdal = TRUE,
    method = "average"
  ) |>
  setNames("infrastructureIndex") |>
  st_transform(myCRS) |>
  mutate(infrastructureIndex = case_when(
    infrastructureIndex < 1 ~ 0,
    infrastructureIndex < 6 ~ 1,
    infrastructureIndex < 12 ~ 2,
    infrastructureIndex >= 12 ~ 3
  )) |>
  # taking away point in the sea
  st_crop(outline)

# This step might seem rather stupid. We want to vectorize a rather large 
# raster. This makes it a quite big data object. The reason is that there is no 
# really good way to burn polygon data on to raster grid cells after the disuse 
# of the raster package. It was not straight forward then either. But 
# calculating intersections between polygons is very fast and easy.

@@

```{r InfraStructureIndex}
#| eval: true

@@

# Calculate area
infra <- infra |>
  mutate(
    area = geometry |> st_area(),
    area_km = area |> set_units("km2")
  )

# show the summered area per HIA
infra |>
  as_tibble() |>
  group_by(infrastructureIndex) |>
  summarise(area_km = sum(area_km)) |>
  ggplot(aes(
    x = infrastructureIndex,
    y = area_km
  )) +
  geom_col()

@@

# Make a plot to check that it has worked
infraMuni3_tbl |>
  ggplot() +
  geom_bar(
    aes(
      x = infrastructureIndex,
      y = total_area_HIAs_km2,
      fill = factor(infrastructureIndex),
      colour = factor(infrastructureIndex)
    ),
    stat = "identity",
    lwd = 1.2
  ) +
  scale_fill_manual(values = RColorBrewer::brewer.pal(4, "YlOrBr")) +
  scale_color_manual(values = RColorBrewer::brewer.pal(5, "YlOrBr")[-1]) +
  theme_minimal_hgrid() +
  labs(
    x = "Human Impact Area (HIA) class",
    y = "Area (km<sup>2</sup>)"
  ) +
  theme(
    axis.title.x = element_textbox_simple(
      width = NULL,
      padding = margin(4, 4, 4, 4),
      margin = margin(4, 0, 0, 0),
      linetype = 1,
      r = grid::unit(8, "pt"),
      fill = "azure1"
    ),
    axis.title.y = element_textbox_simple(
      width = NULL,
      padding = margin(4, 4, 4, 4),
      margin = margin(4, 0, 0, 0),
      linetype = 1,
      orientation = "left-rotated",
      r = grid::unit(8, "pt"),
      fill = "azure1"
    ),
    strip.background = element_blank(),
    strip.text = element_textbox(
      size = 12,
      color = "white", fill = "#5D729D", box.color = "#4A618C",
      halign = 0.5, linetype = 1, r = unit(5, "pt"), width = unit(1, "npc"),
      padding = margin(2, 0, 1, 0), margin = margin(3, 3, 3, 3)
    )
  ) +
  guides(fill = "none", colour = "none") +
  scale_y_log10() +
  facet_grid(cols = vars(Municipality))
```
```{r corrCheck}
#| eval: true
# Now I want to see if the indicator values depend on the HIA is a predictable
# way to justify the stratification
corrCheck <- st_intersection(naturetypes, infra)

# corrCheck |>
#  as_tibble() |>
#  mutate(i_ADSV_fct = floor(round(i_ADSV*10,2))/10,
#         i_alien_fct = floor(round(i_alien*10,2))/10) |>
#  count(i_ADSV_fct, infrastructureIndex) |>
#  View()

@@

```{r HIA-validate}
#| eval: false
# A first look
corrCheck |>
  mutate(
    i_ADSV_fct = floor(round(i_ADSV * 10, 2)) / 10,
    i_alien_fct = floor(round(i_alien * 10, 2)) / 10,
    i_ditch_fct = floor(round(i_ditch * 10, 2)) / 10
  ) |>
  pivot_longer(
    cols = c(i_ADSV_fct, i_alien_fct, i_ditch_fct),
    values_to = "indicatorValue",
    names_to = "indicator",
    values_drop_na = T
  ) |>
  ggplot(aes(
    x = factor(infrastructureIndex),
    fill = factor(indicatorValue)
  )) +
  geom_bar(
    position = "fill"
  ) +
  theme_bw(base_size = 12) +
  guides(fill = guide_legend("Scaled indicator values")) +
  ylab("Fraction of data points") +
  xlab("HIA") +
  scale_fill_brewer(palette = "RdYlGn") +
  facet_grid(indicator ~ year)
```

```{r}
#| eval: false
# The year effect is constant, so ignoring that
corrCheck |>
  mutate(
    i_ADSV_fct = floor(round(i_ADSV * 10, 2)) / 10,
    i_alien_fct = floor(round(i_alien * 10, 2)) / 10,
    i_ditch_fct = floor(round(i_ditch * 10, 2)) / 10
  ) |>
  pivot_longer(
    cols = c(i_ADSV_fct, i_alien_fct, i_ditch_fct),
    values_to = "indicatorValue",
    names_to = "indicator",
    values_drop_na = T
  ) |>
  ggplot(aes(
    x = factor(infrastructureIndex),
    fill = factor(indicatorValue)
  )) +
  geom_bar(
    position = "fill"
  ) +
  theme_bw(base_size = 12) +
  guides(fill = guide_legend("Scaled indicator values")) +
  ylab("Fraction of data points") +
  xlab("HIA") +
  scale_fill_brewer(
    palette = "RdYlGn") +
  facet_grid(~indicator)
```

```{r}
#| eval: true
# Using a color gradient emphasizes the first color (dark green).
# Lets try discrete colors, and merge som classes to simplify
(validationPlot <- corrCheck |>
  #rename(ADSV = i_ADSV) |>
  pivot_longer(
    cols = c(i_ADSV, i_alien, i_ditch),
    values_to = "indicatorValue",
    names_to = "indicator",
    values_drop_na = T
  ) |>
  mutate(
    condition = case_when(
      indicatorValue < 0.6 ~ "<0.6",
      indicatorValue < 0.8 ~ "0.6 to 0.8",
      indicatorValue < 0.91 ~ "0.8 to 0.9",
      .default = "0.9 to 1"
    ),
    condition = fct_reorder(condition, indicatorValue),
    indicator = case_when(
      indicator == "i_ADSV" ~ "ADSV",
      indicator == "i_alien" ~ "Alien species",
      indicator == "i_ditch" ~ "Trenching"
    )
  ) |>
  ggplot(aes(
    x = factor(infrastructureIndex),
    fill = condition
  )) +
  geom_bar(
    position = "fill"
  ) +
  theme_bw(base_size = 12) +
  guides(fill = guide_legend("Indicator values")) +
  ylab("Fraction of data points") +
  xlab("Human Impact Areas") +
  scale_fill_brewer(palette = "RdYlGn") +
  facet_grid(~indicator)
)
```

```{r}
#| eval: false
# The figure changes a lot depending on which value gets the darker colour
corrCheck |>
  pivot_longer(
    cols = c(i_ADSV, i_alien, i_ditch),
    values_to = "indicatorValue",
    names_to = "indicator",
    values_drop_na = T
  ) |>
  mutate(
    condition = case_when(
      indicatorValue < 0.6 ~ "Poor",
      indicatorValue < 0.8 ~ "Intermediate",
      indicatorValue < 0.91 ~ "Good",
      .default = "Perfect"
    ),
    condition = fct_reorder(condition, -indicatorValue)
  ) |>
  ggplot(aes(
    x = factor(infrastructureIndex),
    fill = condition
  )) +
  geom_bar(
    position = "fill"
  ) +
  theme_bw(base_size = 12) +
  guides(fill = guide_legend("Scaled indicator values")) +
  ylab("Fraction of data points") +
  xlab("HIA") +
  scale_fill_viridis_d(alpha = .8) +
  facet_grid(~indicator)
```

```{r}
#| eval: false
# If I look at this on a log scale, it also changes and becomes less convincing for ADSV and alien
corrCheck |>
  pivot_longer(
    cols = c(i_ADSV, i_alien, i_ditch),
    values_to = "indicatorValue",
    names_to = "indicator",
    values_drop_na = T
  ) |>
  mutate(
    condition = case_when(
      indicatorValue < 0.6 ~ "Poor",
      indicatorValue < 0.8 ~ "Intermediate",
      indicatorValue < 0.91 ~ "Good",
      .default = "Perfect"
    ),
    condition = fct_reorder(condition, -indicatorValue)
  ) |>
  ggplot(aes(
    x = factor(infrastructureIndex),
    fill = condition
  )) +
  geom_bar(
    position = "fill"
  ) +
  theme_bw(base_size = 12) +
  guides(fill = guide_legend("Scaled indicator values")) +
  ylab("Fraction of data points") +
  xlab("HIA") +
  scale_y_log10() +
  scale_fill_viridis_d(alpha = .8) +
  facet_grid(~indicator)

@@

# A figure with just the legend
infraMuniMap_l <- tm_shape(muni3) +
  tm_borders() +
  tm_shape(infraMuni3) +
  tm_polygons(
    col = "infrastructureIndex",
    style = "cat",
    title = "Human\nImpact\nArea"
  ) +
  tm_layout(legend.only = TRUE,
            legend.position = c("left", "bottom"),
            legend.outside = TRUE)

# Survey coverage and other colour overlayed the municipalities
muniPlot <- tm_shape(muni3) +
  tm_borders() +
  tm_facets(
    by = "Municipality",
    ncol = 3
  ) +
  tm_shape(terrestrial) +
  tm_fill(
    col = "lightgreen",
    alpha = .4
  ) +
  tm_shape(ocean) +
  tm_polygons(
    col = "skyblue",
    border.col = "black"
  ) +
  tm_shape(dk2) +
  tm_polygons(
    col = "grey",
    alpha = .8
  ) +
  tm_shape(nature3) +
  tm_polygons(
    col = "red",
    border.col = "red",
    lwd = 3
  )
